### 2018-06-08 Writing/response_to_reviewers.tex

\section{Reviewer 1}

\subsection{Major suggestions}

\begin{enumerate}

\item{\textcolor{blue}{I have some concerns about the approach, however. I was somewhat confused by the CD metric. When quantifying n, how is it decided how many taxa have "comparable" characters?
If you simulated the character for all taxa, do they not all have this character (i.e. n always equals the dataset size)?
But more importantly, will two characters not be more similar if they share the same number of character states, independent of if they are telling you the same evolutionary history?
It seems to me that whether or not two characters share a state space is not the relevant axis along which to measure correlation.}}
\label{gower}

Using a Gower based distance, we only compared the comparable taxa between characters.
For example, for two characters \texttt{x = {0, 1, 2, 3}} and \texttt{y = {0, ?, ?, ?}}, the Character Difference metric will be effectively calculated using only the first taxon (\texttt{x = {0}} and \texttt{y = {0}}) and would result in a Character Difference of 0.
As the reviewer points out, characters with less taxa in common will thus be less likely to be different.
This is consistent with what the CD metric actually captures in terms of phylogenetic signal (the number of share splits - see Reviewer 2 comment \ref{proof}).
In the example above, the both characters imply at least one split (between the first taxon and the others).
In other words, in a phylogenetic software, both characters will support the same split.

Besides, as the reviewer also correctly points out, we simulated our data with no missing data so this does not affect our results.
The precision in the manuscript was just describing the properties of the new CD metric.

We've clarified how missing data can be potentially treated in the methods section:
\textit{For example, for two characters X and Y for four taxa, if character \texttt{X = \{0,0,?,1\}} for the first, second, third and fourth taxa respectively and character \texttt{Y = \{0,0,1,1\}}, only the first, second and fourth taxa are compared for this character.
Note that in these simulations, no missing data was included so the characters are compared for equal number of taxa (see below).} lines @@@

@@

\item{\textcolor{blue}{"To assess the effects of character correlation on the accuracy of phylogenetic inference we generated a series of matrices exhibiting different levels of correlation between some characters (Fig.1 - note that each step is described in more details below):"\\
Fig. 1 is the results, but this text implies it is a workflow diagram.}}

We doubled checked the figures links and made sure that this mention of Fig.1 now properly points to the workflow figure.

\item{\textcolor{blue}{"Additionally, we only consider differences for taxa with shared information (i.e. a Gower distance; Gower, 1971)."\\
I'm not sure what this means. Does this mean that you did not include taxa with missing information when calculating per-character distances?}}

We've clarified this sentence (see Reviewer 1, Major suggestion \ref{gower}).

@@

\item{\textcolor{blue}{"The resulting full simulation was 3.5TB big so is not shared here (though the 342 parameters are)."\\
3.5 TB in size.}}

We fixed this typo.

\item{\textcolor{blue}{Most of the time when RF is normalized, it's done so that low scores are better - i.e. a low RF score indicates less distance to the true tree. I was very confused looking at the figures, until I realized this was not how normalizing had been performed. }}

We've changed the labels of the figures to NTS(RF) and NTS(Tr) to avoid any confusions.

### 2018-06-08 Analysis/03-EffectCorrelationFigure.Rmd
Updated figures labels

```{r, fig.width=12, fig.height=12}
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 2, what = "best",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "RF for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 2, what = "best", ylab = "RF for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 2, what = "best",
    ylab = "RF for 150 taxa", NTS = TRUE)
par(op)
```

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE, result = 'hide'}
pdf("../Writing/Figures/RF_results_best.pdf",width = 8,height = 8)
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 2, what = "best",
    legend = FALSE, NTS = TRUE, xlab = "", ylab = "RF for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 2, what = "best", ylab = "RF for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 2, what = "best",
    ylab = "RF for 150 taxa", NTS = TRUE)
par(op)
# dev.off()
```

@@

```{r, fig.width=12, fig.height=12}
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 4, what = "best",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "Tr for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 4, what = "best", ylab = "Tr for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 4, what = "best",
    ylab = "Tr for 150 taxa", NTS = TRUE)
par(op)
```

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE, result = 'hide'}
pdf("../Writing/Figures/Tr_results_best.pdf", width = 8, height = 8)
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 4, what = "best",
    legend = FALSE, NTS = TRUE, xlab = "", ylab = "Tr for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 4, what = "best", ylab = "Tr for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 4, what = "best",
    ylab = "Tr for 150 taxa", NTS = TRUE)
par(op)
# dev.off()
```

@@

```{r, fig.width=12, fig.height=12}
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 2, what = "null",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "RF for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 2, what = "null", ylab = "RF for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 2, what = "null",
    ylab = "RF for 150 taxa", NTS = TRUE)
par(op)
```

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE, result = 'hide'}
pdf("../Writing/Figures/RF_results_null.pdf", width = 8, height = 8)
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 2, what = "null",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "RF for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 2, what = "null", ylab = "RF for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 2, what = "null",
    ylab = "RF for 150 taxa", NTS = TRUE)
par(op)
# dev.off()
```

@@

```{r, fig.width=12, fig.height=12}
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 4, what = "null",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "Tr for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 4, what = "null", ylab = "Tr for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 4, what = "null",
    ylab = "Tr for 150 taxa", NTS = TRUE)
par(op)
```

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE, result = 'hide'}
pdf("../Writing/Figures/Tr_results_null.pdf", width = 8, height = 8)
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 4, what = "null",
    legend = FALSE, NTS = TRUE, xlab = "", ylab = "Tr for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 4, what = "null", ylab = "Tr for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 4, what = "null",
    ylab = "Tr for 150 taxa", NTS = TRUE)
par(op)
# dev.off()
```

>>>>

```{r, fig.width=12, fig.height=12}
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 2, what = "best",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "NTS(RF) for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 2, what = "best", ylab = "NTS(RF) for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 2, what = "best",
    ylab = "NTS(RF) for 150 taxa", NTS = TRUE)
par(op)
```

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE, result = 'hide'}
pdf("../Writing/Figures/RF_results_best.pdf",width = 8,height = 8)
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 2, what = "best",
    legend = FALSE, NTS = TRUE, xlab = "", ylab = "NTS(RF) for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 2, what = "best", ylab = "NTS(RF) for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 2, what = "best",
    ylab = "NTS(RF) for 150 taxa", NTS = TRUE)
par(op)
# dev.off()
```

@@

```{r, fig.width=12, fig.height=12}
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 4, what = "best",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "NTS(Tr) for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 4, what = "best", ylab = "NTS(Tr) for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 4, what = "best",
    ylab = "NTS(Tr) for 150 taxa", NTS = TRUE)
par(op)
```

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE, result = 'hide'}
pdf("../Writing/Figures/Tr_results_best.pdf", width = 8, height = 8)
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 4, what = "best",
    legend = FALSE, NTS = TRUE, xlab = "", ylab = "NTS(Tr) for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 4, what = "best", ylab = "NTS(Tr) for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 4, what = "best",
    ylab = "NTS(Tr) for 150 taxa", NTS = TRUE)
par(op)
# dev.off()
```

@@

```{r, fig.width=12, fig.height=12}
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 2, what = "null",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "NTS(RF) for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 2, what = "null", ylab = "NTS(RF) for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 2, what = "null",
    ylab = "NTS(RF) for 150 taxa", NTS = TRUE)
par(op)
```

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE, result = 'hide'}
pdf("../Writing/Figures/RF_results_null.pdf", width = 8, height = 8)
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 2, what = "null",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "NTS(RF) for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 2, what = "null", ylab = "NTS(RF) for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 2, what = "null",
    ylab = "NTS(RF) for 150 taxa", NTS = TRUE)
par(op)
# dev.off()
```

@@

```{r, fig.width=12, fig.height=12}
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 4, what = "null",
    legend = TRUE, NTS = TRUE, xlab = "", ylab = "NTS(Tr) for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 4, what = "null", ylab = "NTS(Tr) for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 4, what = "null",
    ylab = "NTS(Tr) for 150 taxa", NTS = TRUE)
par(op)
```

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE, result = 'hide'}
pdf("../Writing/Figures/Tr_results_null.pdf", width = 8, height = 8)
op <- par(mfrow=c(3,1), mar = c(2, 4, 1, 2))
plot.results.single(t25_list, metric = 4, what = "null",
    legend = FALSE, NTS = TRUE, xlab = "", ylab = "NTS(Tr) for 25 taxa",
    axislab = FALSE)
plot.results.single(t75_list, metric = 4, what = "null", ylab = "NTS(Tr) for 75 taxa",
    xlab = "", axislab = FALSE, NTS = TRUE)
plot.results.single(t150_list, metric = 4, what = "null",
    ylab = "NTS(Tr) for 150 taxa", NTS = TRUE)
par(op)
# dev.off()
```

### 2018-06-08 Writing/draft_highlighted.tex
Updated figures labels

\noindent Of course, the three sources of dependence have an interaction: characters describing the left and right lower/upper molars will have induced dependence due to the modularity of the molars, their shared history and the duplicated coding.
Logical dependence, however, is easily distinguished prior to phylogenetic inference, while the two other ones (intra-organismal and evolutionary) are much harder.
However, the development of algorithms and software has not yet caught up with the need to deal with these interdependencies \citep{DeLaet2015,BrazeauNA}.
Intra-organismal dependence requires more detailed, often extremely time-consuming studies (and possibly beyond the limits of available technology).
Even after all of the effort is expended, the results might then only be known for a single (model) species.
Evolutionary dependence itself requires the resolution of a phylogenetic tree, and is best determined by independent character sets. 
This is frequently accomplished by mapping morphological traits on molecular phylogenetic trees.

@@

\subsection{Measuring differences between characters}
\label{CD_description}
To measure the effect of character correlation as interpreted by the phylogenetic software, we define characters as being entirely correlated if they give the same phylogenetic information.
In order to measure this, we propose a new distance metric to measure the difference between two characters:
%TG: OK, I've checked Farris' stuff. Despite not beeing described (at least not the in 1970 Syst Biol paper), it seems that the advancement index is just the distance between two taxa. Appart from the taxa/character difference, it's kind of similar but (1) it's not ponderated (divided by n), centered (-1/2) and reversed (1-2(...)); so I'm not sure if we need to say it's similar. In the end it's a distance metric and it can be pretty similar to *any* distance metric. I think we can leave as it is (plus, I've reinvented the wheel by mathematicallty demonstrating it's a distance - my first only real math in a draft).

@@

\noindent Where $n$ is the number of taxa with comparable characters $x$, $y$ and $x_i$, $y_i$ are each character's state for the $i^{th}$ taxon.
$CD$ is a continuous distance metric bounded between $0$ and $1$ (see the mathematical demonstration in the supplementary material 1).
Since we are considering differences as being only Fitch-like (non-additive) and unweighted, we calculated the difference between character states in a qualitative way.
Two same character states tokens have a difference of zero and two different ones have a difference of one (e.g. $0 - 0 = 0$ or $1 - 8 = 1$).
Additionally, we only consider differences for taxa with shared information \citep[i.e. a Gower distance;][]{GowerDist}.
\hl{For example, for two characters X and Y for four taxa, if character \texttt{X = \{0,0,?,1\}} for the first, second, third and fourth taxa respectively and character \texttt{Y = \{0,0,1,1\}}, only the first, second and fourth taxa are compared for this character.}

@@

\subsection{Effect of character differences on topology}
As expected, there is a significant effect of the character difference in the ability to recover the correct topology.
The character difference metric can be seen as the inverse of character correlation (see Methods \ref{CD_description}): a high character difference approximates a low level of character correlation and vice versa.
When characters are correlated, one could expect the matrices to convey a strong (but potentially misleading) phylogenetic signal since every character agrees with each other and conversely, when characters are uncorrelated, one could expect them to convey a weaker phylogenetic signal with a high amount of homoplasy.
Intuitively, this would lead the ``minimised'' character difference scenario to lead to incorrect but consistent trees, the ``maximised'' scenario to lead to poorly resolved once (really homoplasic trees) and the ``randomised'' scenario to perform the best at recovering the correct topology.
Although the expected results appear to be true for a low character difference scenario, increasing the character difference surprisingly improves the ability to recover the ``normal'' topology both in terms of clade conservation ($NTS_{RF}$) and taxa placement ($NTS_{Tr}$) for both inference methods (especially in bigger matrices; Figs \ref{Fig:RF_results_best} and \ref{Fig:Tr_results_best}).
Furthermore, the trees generated by the ``minimised'' scenario do not appear better resolved (towards any topology) than the other scenarios (see Supplementary material 3, Figs 3, 4 and 5).

@@

\subsection{Conclusion}
Correlation between characters can be induced through three main phenomena: intra-organismal relationships, selection-driven covariation or biases in coding the characters yet only the latter can be improved upon to investigate phylogenetic relationships.
Useful best practices guidelines \citep[e.g.][]{Brazeau2011,simoes2017giant} and algorithms for dealing with different types of character correlations \citep[e.g. for characters hierarchy][]{deLaet2015,BrazeauNA} already exist.
However, with the regain of popularity in discrete morphological data and the expansion of dataset size \citep[e.g.][with more than 1000 characters each]{nithe2013,O'Leary08022013}, we can expect the correlation between characters to increase stochastically.
Moreover, because phylogenetic inference software are unable to \textit{a priori} differentiate these difference correlations, it is important to understand to what extant topologies can be induced by such bias.

@@

\section{Acknowledgements}
Calculations where done using the Imperial College London Cluster Services (doi: 10.14469/hpc/2232).
We thank Alberto Pascual Garcia for input in the design of the simulations protocol and Guillermo Herraiz Yebes for helping with the CD metric proof.

>>>>

\noindent Of course, the three sources of dependence have an interaction: characters describing the left and right lower/upper molars will have induced dependence due to the modularity of the molars, their shared history and the duplicated coding.
Logical dependence, however, is easily distinguished prior to phylogenetic inference, while the two other ones (intra-organismal and evolutionary) are much harder.
However, the development of algorithms and software has not yet caught up with the need to deal with these interdependencies \citep{de2015parsimony,BrazeauNA}.
Intra-organismal dependence requires more detailed, often extremely time-consuming studies (and possibly beyond the limits of available technology).
Even after all of the effort is expended, the results might then only be known for a single (model) species.
Evolutionary dependence itself requires the resolution of a phylogenetic tree, and is best determined by independent character sets. 
This is frequently accomplished by mapping morphological traits on molecular phylogenetic trees.

@@

\subsection{Measuring differences between characters}
\label{CDdescription}
To measure the effect of character correlation as interpreted by the phylogenetic software, we define characters as being entirely correlated if they give the same phylogenetic information.
In order to measure this, we propose a new distance metric to measure the difference between two characters:

@@

\noindent Where $n$ is the number of taxa with comparable characters $x$, $y$ and $x_i$, $y_i$ are each character's state for the $i^{th}$ taxon.
$CD$ is a continuous distance metric bounded between $0$ and $1$ (see the mathematical demonstration in the supplementary material 1).
Since we are considering differences as being only Fitch-like (non-additive) and unweighted, we calculated the difference between character states in a qualitative way.
Two same character states tokens have a difference of zero and two different ones have a difference of one (e.g. $0 - 0 = 0$ or $1 - 8 = 1$).
Additionally, we only consider differences for taxa with shared information \citep[i.e. a Gower distance;][]{GowerDist}.
\hl{For example, for two characters X and Y for four taxa, if character \texttt{X = \{0,0,?,1\}} for the first, second, third and fourth taxa respectively and character \texttt{Y = \{0,0,1,1\}}, only the first, second and fourth taxa are compared for this character.
Note that in these simulations, no missing data was included so the characters are compared for equal number of taxa (see below).}

@@

\subsection{Effect of character differences on topology}
As expected, there is a significant effect of the character difference in the ability to recover the correct topology.
The character difference metric can be seen as the inverse of character correlation (see Methods above): a high character difference approximates a low level of character correlation and vice versa.
When characters are correlated, one could expect the matrices to convey a strong (but potentially misleading) phylogenetic signal since every character agrees with each other and conversely, when characters are uncorrelated, one could expect them to convey a weaker phylogenetic signal with a high amount of homoplasy.
Intuitively, this would lead the ``minimised'' character difference scenario to lead to incorrect but consistent trees, the ``maximised'' scenario to lead to poorly resolved once (really homoplasic trees) and the ``randomised'' scenario to perform the best at recovering the correct topology.
Although the expected results appear to be true for a low character difference scenario, increasing the character difference surprisingly improves the ability to recover the ``normal'' topology both in terms of clade conservation ($NTS_{RF}$) and taxa placement ($NTS_{Tr}$) for both inference methods (especially in bigger matrices; Figs \ref{Fig:RF_results_best} and \ref{Fig:Tr_results_best}).
Furthermore, the trees generated by the ``minimised'' scenario do not appear better resolved (towards any topology) than the other scenarios (see Supplementary material 3, Figs 3, 4 and 5).

@@

\subsection{Conclusion}
Correlation between characters can be induced through three main phenomena: intra-organismal relationships, selection-driven covariation or biases in coding the characters yet only the latter can be improved upon to investigate phylogenetic relationships.
Useful best practices guidelines \citep[e.g.][]{Brazeau2011,simoes2017giant} and algorithms for dealing with different types of character correlations \citep[e.g. for characters hierarchy][]{de2015parsimony,BrazeauNA} already exist.
However, with the regain of popularity in discrete morphological data and the expansion of dataset size \citep[e.g.][with more than 1000 characters each]{nithe2013,O'Leary08022013}, we can expect the correlation between characters to increase stochastically.
Moreover, because phylogenetic inference software are unable to \textit{a priori} differentiate these difference correlations, it is important to understand to what extant topologies can be induced by such bias.

@@

\section{Acknowledgements}
Calculations where done using the Imperial College London Cluster Services (doi: 10.14469/hpc/2232).
We thank Alberto Pascual Garcia for input in the design of the simulations protocol and Guillermo Herraiz Yebes and Abigail Pastore for helping with the CD metric proof and the \hl{three anonymous reviewers for there insightful comments}.
